---
title: "Cleaning data using Rmarkdown"
author: "Huruy Belay"
date: "May 19, 2017"
output: html_document
---

## Data cleaning on school attendance

We will work with attendance data from public schools in the US, organized by school level and state, during the 2007-2008 academic year. The data contain information on average daily attendance (ADA) as a percentage of total enrollment, school day length, and school year length.

I got the data from Datacamp training.

We can [view](http://s3.amazonaws.com/assets.datacamp.com/production/course_1294/datasets/attendance_screenshot.png) a screenshot of the data.

#### Loading libraries

```{r, results='hide',warning=FALSE, message=FALSE}
library(gdata)
library(stringr)
library(dplyr)
```

#### Importing data

In order to inmport the xls file, we have to download perl first.

```{r, results='hide'}
att <- read.xls("attendance.xls", perl = "C:\\Perl\\bin\\perl.exe")
att
```

#### Examining the structure 

The next step should be examining the data through the following:

```{r, results='hide'}
names(att)
head(att)
tail(att)
str(att)
```

#### Removing unnecessary rows and columns

* **Remove rows 3 and from 66 to 69**

```{r}
att2 <- att[-c(3, 66:69), ]
```

* **Remove columns 3, 5, 7, 9, 11, 13, 15, 17 **

```{r}
att3 <- att2[, -c(3, 5, 7, 9, 11, 13, 15, 17)]
```

#### Splitting the data

If you observe the dataset stores multiple tables of information. You can often diagnose this problem by looking at the column names and noticing duplicate rows.

* Columns 1, 6, and 7 represent attendance data for US elementary schools,
* Columns 1, 8, and 9 represent data for secondary schools and
* columns 1 through 5 represent data for all schools in the US.

Each of these should be stored as its own separate data frame, therefore we should split them.

* **Subset for elementary schools:**

```{r}
elem_att <- att3[, c(1, 6, 7)]
```

* ** Subset for secondary schools:**

```{r}
seco_att <- att3[, c(1, 8, 9)]
```

* **Subset for all schools:**

```{r}
att4 <- att3[, 1:5]
```

#### Replacing the names

It is better if we modified the column names:

```{r}
cnames <- c("state", "avg_attend_pct", "avg_hr_per_day", 
            "avg_day_per_yr", "avg_hr_per_yr")

colnames(att4) <- cnames

```

Nice and better if we remove the first two rows of all_att data.

```{r}
att5 <- att4[-c(1, 2), ]
```

#### Cleaning up extra characters

If we look the variable state on the att5 dataset, the state variable should have to clean. To clean the variable we will use the regular expression. 

```{r}
# Remove all periods in state column

att5$state <- str_replace_all(att5$state, "\\.", "")
```

```{r}
# Remove white space around state names

att5$state <- str_trim(att5$state)
```

```{r}
# View the head of att5

head(att5)
str(att5)
```

Finally we should have to change some of the variables to numeric. We can change by two methods.

```{r}
# Using the dply package:
att5 <- mutate_each(att5, funs(as.numeric), -state)
```

**OR**

```{r}
# using the sapply 
att5[, 2:ncol(att5)] <- sapply(att5[, 2:ncol(att5)], as.numeric)
```

** Checking using str function**

```{r}
str(att5)
```

